A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-15 paper/paper.tex
Minor tweaks

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed lambda calculus and System F, a streamlined
presentation of proof by logical relation for a dependently language
is lacking in comparison. In this paper, I present a short
consistency proof for a dependently typed language that contains a
rich set of features, including an infinite cumulative universe
hierarchy, booleans, and an intensional identity type. We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code.
\end{abstract}

@@

Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages, and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity. \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. In short, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms.

The glaring gap here is the
lack of fully dependently typed systems where computations may appear
at the type-level. It is far from obvious why proof by
logical relation is even applicable to dependent types, since the type
may very well-be a computation that is yet to be evaluated.
While it is assuring that proof by logical relations for dependent
types is available in mechanized forms, % the key to address the
% complexities of dependent types is obscured in
\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
Kripke-style models that obscure the technique for addressing
type-level computation. The added complexivity is evident
from the size of their code base. For example, \citet{nbeincoq}
involves 22 thousand lines of Coq code.

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format. % The language specification will
% be given in Section~\ref{sec:spec}.
The system we present is small but not minimal. We want to cover
features that are only available or require special treatments under
dependent types%  without introducing complexities that would have
% existed in its simply typed variant
. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system as the interpretation
of identity types requires an extra side condition that is needed for
indexed family of types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
but similar to \citet{anand2014towards}, we include an infinite
hierarchy of universes since we want to not only have at least one
predicative universe level to support type-level computations, but
also avoid the unnecessary code duplication pointed out by
\citet{nbeincoq}.

Our consistency proof is short and fully mechanized. The proof scripts
involve less than 1000 lines of Coq code. In fact, what is even more
encouraging is that among the 1000
lines of Coq code, 400 lines are related to the specification of the
type system, semantics, and properties related to untyped lambda
terms. The semantic type soundness proof through logical relation
takes almost the same amount of code as our syntactic type soundness
proof!
Thanks to the conciseness of
the proof, we are able to present most of its details in
Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
structure of our mechanization closely corresponds to the proof we
present in the text, enabling us to label at the footnote each lemma
directly to their counterpart in the proof script for the readers to
reference and validate.

The technique we use is most similar to the one from \citet{nbeincoq},
which leverages impredicativity to define the logical
relation. % In our presented proof, we use the informal
% language of sets, where the use of impredicativity feels more natural
% and less noticeable.
Rather than framing impredicativity as a
mechanism for encoding induction
recursion~\citep{induction-recursion-dybjer}, a scheme in which
semantic models (including logical relations) for dependent types can
be defined, we opt for a direct explanation in informal language of
sets, where impredicativity manifests in the form of second-order
logical formulas and thus more intuitive to understand for readers
with general mathematics background.

@@

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation.

Finally, in Section~\ref{sec:relatedwork}, we discuss how our proof
relates to existing proofs by logical relations for dependent types in
the literature.

@@

We sketch out some key properties about parallel reduction and
coherence without giving their proofs. Our technique for establishing
those results is based on \citet{takahashi-parallel-reduction} and a
modern exposition of the same technique can be found in
\citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}

@@

% From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
% single substitution property as a simple corollary.
% \begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
%   \label{lemma:parsubst}
%   If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
% \end{corollary}

@@

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity described
in \citet{siles2012pure}. Section~\ref{sec:extension} explains how we
generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

@@

\section{Logical Relation}
\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}). The
logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures:
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of subsets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.

@@

By unfolding the definition from Figure~\ref{fig:logrelrec}, we can
now show that the same introduction rules for $[[Interp I i A S]]$ are
admissible for $[[InterpR i A S]]$, by
instantiating $[[I]]$ with $[[I i]] := [[{A | exists S , InterpR i A
  S}]]$, the same function $[[I]]$ used in the definition of $[[InterpR i A
S]]$. We give as example the following rules.
\begin{center}
\drule[]{IR-Void} \qquad \drule[]{IR-Set}
\end{center}

@@

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
witness in the precondition for $[[a in S]]$.
With the alternative representation that is free of the function $[[F]]$, we
might want to reformulate our logical relation in the following way.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction
principle. However, as we will see in Lemma~\ref{lemma:piintroalt} to
show the admissibility of \rref{I-Pi} a posteriori.

@@

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We leave features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts as part of our future work.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently language
is lacking. In this paper, I present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a full cumulative universe
hierarchy, booleans, and an intensional identity type. We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code.
\end{abstract}

@@

Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages, and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity. \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. Overall, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms.

The glaring gap here is the
lack of fully dependently typed systems where computations may appear
at the type-level. It is far from obvious why proof by
logical relation is even applicable to dependent types, since the type
may very well-be a computation that is yet to be evaluated.
While it is assuring that proof by logical relations for dependent
types is available in mechanized forms, % the key to address the
% complexities of dependent types is obscured in
\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
Kripke-style models that obscure the technique for addressing
type-level computation. The added complexivity is evident
from the size of their code base. All three developments involve 20
thousand to 30 thousand lines of Coq or Agda code.

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format.
The system we present is small but covers
important features that are only available or require special treatments in
dependently typed systems. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system to show how indexed
types are treated in a logical relational proof. Unlike
\citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
\citet{anand2014towards}, we include an infinite hierarchy of
universes to not only support type-level computations, but also avoid
the unnecessary code duplication pointed out by \citet{nbeincoq} when
big and small types are treated non-uniformly.

Our consistency proof is short and fully mechanized. The proof scripts
involve less than 1000 lines of manually written Coq code. In fact, what we find
encouraging is that among the 1000
lines of Coq code, 400 lines are related to the specification of the
type system, semantics, and properties related to untyped lambda
terms. The semantic type soundness proof through logical relation
takes almost the same amount of code as our syntactic type soundness
proof!
Thanks to the conciseness of
the proof, we are able to present it in detail in
Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
structure of our mechanization closely corresponds to the proof we
present in the text, enabling us to label at the footnote each lemma
directly to their counterpart in the proof script for the readers to
reference and validate.

The technique we use is most similar to the one from \citet{nbeincoq},
which leverages impredicativity to define the logical
relation as a partial function.
Rather than framing impredicativity as a
mechanism for encoding induction
recursion~\citep{induction-recursion-dybjer}, a scheme in which
semantic models (including logical relations) for dependent types can
be defined, we opt for a direct explanation through the informal
language of sets, where impredicativity manifests in the form of
second-order logical formulas and thus more intuitive to grasp
for readers with a general mathematical background.

@@

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation.
In Section~\ref{sec:relatedwork}, we discuss how our proof
relates to existing proofs by logical relations for dependent types in
the literature.

@@

We sketch out some key properties about parallel reduction and
coherence without giving their proofs. Our technique for establishing
those results is based on \citet{takahashi-parallel-reduction} and a
modern exposition of the same technique can be found in
\citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}

@@

% From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
% single substitution property as a simple corollary.

@@

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity.
% described
% in \citet{siles2012pure}
Section~\ref{sec:extension} explains how we
generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

@@

\section{Logical Relation}
\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}). The
logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures:
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.

@@

With the definition from Figure~\ref{fig:logrelrec}, we can
show that the same introduction rules for $[[Interp I i A S]]$ are
admissible for $[[InterpR i A S]]$, by
instantiating $[[I]]$ with $[[I i]] := [[{A | exists S , InterpR i A
  S}]]$, the same function $[[I]]$ used in the definition of $[[InterpR i A
S]]$. We give as example the following rules.
\begin{center}
\drule[]{IR-Void} \qquad \drule[]{IR-Set}
\end{center}

@@

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
witness in the precondition for $[[a in S]]$.
With the alternative representation that is free of the function $[[F]]$, we
might want to reformulate our logical relation in the following way.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction
principle. However, as we will see in Lemma~\ref{lemma:piintroalt}, it
is possible to show the admissibility of \rref{I-PiAlt} a posteriori.

@@

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We hope features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts can be implemented in a similarly streamlined
fashion and we will leave those as part of our future work.
